---
title: "CHM NDVI filter"
output:
  pdf_document: default
  html_notebook: default
  word_document: default
Author: Hugo Hunt
Date of last revision: 05/08/2024
---

Firstly, clear the work environment.
```{r}
rm(list=ls())
wd <- getwd()
```
The working directory is by default the location of this markdown file, so does not need to be changed. 

Import the required libraries for this tool
```{r}
library(terra)
library(ggplot2)
library(gdalraster)
library(sf)
library(tmap)
library(rgl)
library(raster)
library(lidR)
library(readxl)
library(magick)
library(ggplot2)
library(gifski)
library(magick)
library(RColorBrewer)
library(dplyr)
library(reshape2)
```

Define locations and file names of the downloaded LiDAR files and Infrared Imagery.
Import the DSM and DTM files as rasters using the rast() function. Calculate the chm by subtracting dtm from dsm
```{r}
dsm_file <- paste0(wd,"/data/NY30NELIDAR/england-dsm-fr-1m_5500023/ny/ny30ne_fz_dsm_1m.tif")
dtm_file <- paste0(wd,"/data/NY30NELIDAR/england-dtm-1m_5500024/ny/ny30ne_dtm_1m.tif")
ir_file <- paste0(wd, "/data/NY30NELIDAR/IR/NY30neIR.tif")

dsm <- rast(dsm_file)
dtm <- rast(dtm_file)
chm <- dsm - dtm
```


Create a new variable, "chmpos". Define it as all the non-negative canopy heights. I.e all canopy heights less than 0 are set to 0.
```{r}
chmpos<-chm
chmpos[chm<0]<-0
plot(chmpos, main="NY30ne non-negative CHM",breaks = c(1, 3, 16, 48))
#hist(chmpos[],main=NULL, breaks = 20, xlim = c(3, 40), ylim = c(0, 1000000))
```
Optionally, the CHM can be plotted over a hillshade of the terrain (DTM). The Canopy heights are classified into scrub (1-3m) and trees (>3m)
```{r}
slope <- terrain(dtm, "slope", unit="radians")
aspect <- terrain(dtm, "aspect", unit="radians")
hill <- shade(slope, aspect, 45, 270)
plot(hill, col=grey(0:100/100), legend=FALSE, mar=c(2,2,2,2), xlim = c(337900, 338200), ylim = c(507700, 508000), main="NY30ne DSM Hillshade with CHM overlaid")
plot(chmpos, col=rainbow(10,alpha=1,rev=TRUE, end = 1- max(1, 10 - 1)/10, start = 0.6), add=TRUE, breaks = c(1, 3, 48))
```

Rasterise the Near-Infrared band (the first band/layer) and the Red band (second band/layer) of the Colour IR data
Since the IR data has a 50cm resolution, decrease resolution from 50cm by factor of 2 to match LiDAR data resoltion of 1m, using the aggregate function. This is less computationally expensive than continuing the analysis and aggregating later in the process.
We can then plot them, with a custom monochrome colour scale for the Red. 
```{r}
IR50 <- rast(ir_file, lyrs=1)
Red <- rast(ir_file,lyrs=2)
IR1 <- aggregate(IR50, fact=2, fun=mean, expand=TRUE)
Red1<-aggregate(Red, fact=2, fun=mean, expand=TRUE)

Redcolours <- colorRampPalette(c("#FFFFFF80", "#FF000080"))(100)
plot(hill, col=grey(0:100/100), legend=FALSE, main="NY30ne Red", mar=c(2,2,2,2))
plot(Red1, col=Redcolours, add=TRUE)

plot(hill, col=grey(0:100/100), legend=FALSE, main="NY30ne IR", mar=c(2,2,2,2))
plot(IR1, col=rainbow(50,alpha=0.5, start=0.2), add=TRUE)
```

Create a function to determine NDVI fora given input of near infrared and red band data
```{r}
calculate_ndvi <- function(nir, red) {
  ndvi <- (nir - red) / (nir + red)
  return(ndvi)
}
```

Calculate the NDVI for NY30ne using the aggregated data
```{r}
ndvi <- calculate_ndvi(IR1, Red1)
plot(hill, col=grey(0:100/100), legend=FALSE, main="NY30ne NDVI", mar=c(2,2,2,2))
plot(ndvi, col=rainbow(50,alpha=0.5, start=0.2), add=TRUE)
```

Input desired threshold value
```{r}
threshold <- 0.18
chmfilt<-chmpos
chmfilt[ndvi < threshold] <- 0

```

```{r}
xfocus <- 335400 #337500
yfocus <- 506800 #506000
radius <- 200
xlower <- xfocus - radius
xupper <- xfocus + radius
ylower <- yfocus - radius
yupper <- yfocus + radius

```

To help visualise effect of NDVI Threshold value, we can plot the CHM for for the current threshold and unfiltered.
```{r}
  #first plot the DTM hillshade, with optional limits
  plot(hill, main=paste("Unfiltered CHM"), 
       #plot(hill, main=paste("Filtered CHM with threshold",threshold),
       #optional defining of axes to focus on a specific part of the tile
       xlim = c(xlower, xupper), ylim = c(ylower, yupper), 
       col=grey(0:100/100), legend=FALSE, mar=c(2,2,2,2))
  #Then plot CHM, with no transparency and the same classification as previously 
  plot(chmpos, col=rainbow(10,alpha=1,rev=TRUE, end = 1- max(1, 10 - 1)/10, start = 0.6), add = TRUE, breaks = c(1, 3, 48))
  
  plot(hill, main=paste("Filtered CHM with threshold",threshold),
       #plot(hill, main=paste("Filtered CHM with threshold",threshold),
       #optional defining of axes to focus on a specific part of the tile
       xlim = c(xlower, xupper), ylim = c(ylower, yupper),  
       col=grey(0:100/100), legend=FALSE, mar=c(2,2,2,2))
  #Then plot CHM, with no transparency and the same classification as previously 
  plot(chmfilt, col=rainbow(10,alpha=1,rev=TRUE, end = 1- max(1, 10 - 1)/10, start = 0.6), add = TRUE, breaks = c(1, 3, 48))

```

