---
title: "PointCloud Processing Notebook"
output:
  word_document: default
  html_notebook: default
date: "05/08/2024"
---

clear workspace and load libraries
```{r Libraries}
rm(list=ls())
library(lidR)
library(ggplot2)
library(RCSF)
library(gstat)
library(rgl)
library(magick)
```

Read in specified LAS file data. Note, a 5x5km tile is very heavy and may take significant time to process. Depending on the workstation it may crash R when plotting. For illustration, a 1x1km tile is used instead. 
```{r LAS_import}
#specify point cloud filename
laz_file <- "data/england-laz-2009_5499952/ny/ny3506_p_6854_20091208_20091218.laz"
#read laz file, filtering for first returns only.
lasfz <- readLAS(laz_file,filter= "-keep_first")
```

Now we can plot the pointcloud in 3d. We can interact with the plot usingf the mouse, or set it to spin at a defined speedm which we then save to a specified directory within the Wd.
```{r 3Dplot}
p <- plot(lasfz, bg="black", axis = TRUE, legend = TRUE, breaks = "quantile")
#play3d(spin3d())
directory<-"GIF"
#movie3d(spin3d(axis = c(0,0,1), rpm = 8), duration = 8, fps = 20, dir = output_dir,movie = "NabScarSpin")
```

Here we can set another spinning 3D plot, but zoomed in with a slow rotation so that individual points can be seen in contenxt
```{r Demo}
p <- plot(lasfz, bg="black", axis = TRUE, legend = TRUE, breaks = "quantile")
rgl.viewpoint(theta = 0, phi = -90, fov = 40, zoom = .1)
#play3d(spin3d(axis = c(0,0,1), rpm = 0.25))
```

Performing further analysis, we can classify the vegetation using the lidR package at the filter_poi function

```{r veg_classification}
veg <- filter_poi(lasfz, Classification == LASHIGHVEGETATION)
nonveg <- filter_poi(lasfz, Classification != LASHIGHVEGETATION)
```

Now we can plot the vegetation and non-vegetation points in different colours to differentiate them easily
```{r PLotting_veg}
x <- plot(nonveg, 
          #color = "Classification", 
          bg = "white", size = 1)
plot(veg, add = x)
```
Alternative classifications
```{r PMFclassification}
las <- readLAS(laz_file, select = "xyzrn")
las <- classify_ground(las, algorithm = pmf(ws = 5, th = 3))
plot(las, color = "Classification", size = 1.5, bg = "black") 
play3d(spin3d())
```
Cloth draping method of classication
```{r ClothDraping}
las <- readLAS(laz_file, select = "xyzrn")
las <- classify_ground(las, algorithm = csf())
plot(las, color = "Classification", size = 1, bg = "white")
```
Rasterise to create a DTM. First using TIN method
```{r TINDTM}
dtm_tin <- rasterize_terrain(las, res = 1, algorithm = tin())
plot_dtm3d(dtm_tin, bg = "white") 
```
Finally, using the KNN IDW (K-nearest neighbour, Inverse Distance Weighting) method
```{r NDDIDW_DTM}
dtm_idw <- rasterize_terrain(las, 1, algorithm = knnidw(k = 10L, p = 2))
d <- plot_dtm3d(dtm_idw, bg = "white", legend = TRUE) 
```

Finally, we can plot the DTM and overlay vegeation
```{r DTMplot}
v <- plot(veg, 
          #color = "Classification", 
          bg = "white", size = 1)
add_dtm3d(v, dtm_idw)
#play3d(spin3d(axis = c(0,0,1), rpm = 4))
```

